---
title: "Interactive Visualization for Slide Presentation"
author: "Cheng Peng"
date: "3/22/2022"
output: slidy_presentation
---

<style type="text/css">
body {
    color: blue;
   }

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 22px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}
h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}
</style>


```{r setup, include=FALSE}
if (!require("tidyverse")) {
   install.packages("tidyverse")
   library(tidyverse)
}
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("cowplot")) {
   install.packages("cowplot")
   library(cowplot)
}
if (!require("latex2exp")) {
   install.packages("latex2exp")
   library(latex2exp)
}
if (!require("plotly")) {
   install.packages("plotly")
   library(plotly)
}
if (!require("gapminder")) {
   install.packages("gapminder")
   library(gapminder)
}
if (!require("png")) {
    install.packages("png")             # Install png package
    library("png")
}
if (!require("RCurl")) {
    install.packages("RCurl")             # Install RCurl package
    library("RCurl")
}
if (!require("colourpicker")) {
    install.packages("colourpicker")              
    library("colourpicker")
}
if (!require("gganimate")) {
    install.packages("gganimate")              
    library("gganimate")
}
if (!require("gifski")) {
    install.packages("gifski")              
    library("gifski")
}
if (!require("magick")) {
    install.packages("magick")              
    library("magick")
}
if (!require("grDevices")) {
    install.packages("grDevices")              
    library("grDevices")
}
if (!require("jpeg")) {
    install.packages("jpeg")              
    library("jpeg")
}
knitr::opts_chunk$set(echo = FALSE,       
                      warning = FALSE,   
                      result = TRUE,   
                      message = FALSE)
```

# Slide Title

The number of hash-tags represents the level of the header. we can align these headers using CSS. For example, the CSS styles that used in creating my class notes.




* You can include any images (including interactive images) in the HTML slide.

```{r}
## preparing data
poc <- read.csv("https://github.com/pengdsci/datasets/raw/main/POC.csv")[,c(7,8,9, 17)]
poc.site <- poc[poc$POC == 1,]
# geo styling
geostyle <- list(scope = 'usa',
                 projection = list(type = 'albers usa'),
                 showland = TRUE,
                 landcolor = toRGB("gray95"),
                 subunitcolor = toRGB("gray85"),
                 countrycolor = toRGB("gray85"),
                 countrywidth = 0.5,
                 subunitwidth = 0.5
               )
## plotting map
fig <- plot_geo(poc.site, lat = ~ycoord, lon = ~xcoord) %>%
       add_markers(text = ~ SITE_DESCRIPTION, 
                   color = "red", 
                   symbol = I("circle"), 
                   size = I(8), 
                   hoverinfo = "text" )   %>%
        layout( title = 'POC Risk Sites', geo = geostyle)
fig
```

# External Images

* You can include static images generated by other device into the HTML slide presentation through R Markdown. 

* Including an interactive images generated by other devices into R markdown may not be straightforward. 

* We can include animated gif images (it is not considered as an interactive image) in R markdown as usual.


# Including Tableau Chars

We cannot include a interactive plot hosted on a web server through R Markdown! One way to get this around is to add the code provided by Tableau into the knitted HTML directly.







